---
title: "R fundamentals"
teaching: 
exercises: 
questions:
- "What are the basic features of R?"
objectives:
- "Start to get to grips with R coding"
keypoints:
- "Base R features and techniques"
---

### Getting help

The easiest way to get help is to write the name of the function you're interested in, preceded by a question mark. For example,

```{r}
#?plot
```

This brings up the help file for the function (a function is a package of code that performs a particular task). Beyond this, there are many websites with guides and user forums with guidance. Often, just searching on Google for the problem you're trying to solve (e.g. *'How do I plot a histogram in R?'*) or searching for the error message (e.g. *'non-numeric argument to binary operator'*) will lead you to the answer.

There are also a number of websites that may be of use, including,

[The Comprehensive R Archive Network](https://cran.r-project.org/)

[Stack Overflow](https://stackoverflow.com/questions/tagged/r)

[Stat Methods - Quick R](https://www.statmethods.net/)

#### Exercise

1. Type and run,

```{r}
mean(numbers)
```

What happens?

2. Type and run,

```{r}
numbers = c(1,2,3,4,NA)
```

What happens?

3. Re-run the following line,

```{r}
mean(numbers)
```

Now what happens?

4. Type '?mean'. See if you can figure out how to amend the code in step 3 to give you a sensible answer

You may notice RStudio's auto-complete feature when typing the answer.


### Simple arithmetic

R is basically a very fancy calculator. You can use arithmetic operators as you would expect, such as,

**addition**

```{r}
3+4
```

**Exponentiation**

```{r}
2^3
```

Use of **brackets**

```{r}
(5+5) / 5
```

### Comparisons

Use greater than/less than symbols such as,

```{r}
3 > 4
```

and,

```{r}
5 == (2.5 * 2)
```

Where **==** means 'equal to'. So, the above command is says *5 is equal to 5*, which then evaluates to *true.*


### Assigning variables

We can give variables names and then consequently use them,

```{r}
sample1 = 6
sample2 = 4
sample1 + sample2
```

We can also assign words to variables,

```{r}
patient_name = "Jim"
```

If we then call that variable, the data we've stored will be printed to the screen,

```{r}
patient_name
```

If we try to mix types, we'll get an error,

```{r}
patient_name + sample1
```

Why? Because we're trying to add a number to a word!

**Vectors**

Vector is just a fancy word for a collection of items. For example,

```{r}
patients_ages = c(12,5,6,3,7,13)
patients_names = c('Jim', 'John', 'Brian', 'Susan', 'Keith', 'Geoff')
```
  
Note that we string them together the the letter *c*, which stands for **combine**.

Once we have a vector, we can pick out certain elements. For example,

```{r}
patients_names[3]
```

Use a colon to indicate items between two indices,

```{r}
patients_names[3:5]
```

Remember using *==* above? Let's try it here,

```{r}
patients_names == 'Jim'
```

We've just said, 'do the elements of our vector, patient_names, equal Jim?'

We can do something similar to the numeric vector,

```{r}
patients_ages >= 10
```

We've just said 'are the elements of our vector, patients_ages, greater than or equal to 10?'

### Built-in Functions

R comes with a huge number of functions, covering every possible data analysis and statistical requirement.

#### Exercise

Create the following vector,

```{r}
numbers = c(2.34, 4.53, 1.42, 1.42, 9.41, 8.54)
```

and investigate the following functions,

- round()
- sqrt()
- mean()
- median()
- max()
- sort()

These are useful functions for applying to existing numerical data. There are also functions that are useful for being applied to non-numerical data.

#### Exercise

Create the following vector,

```{r}
terms = c('Serum', 'Blood', 'Plasma', 'Blood', 'Blood', 'Plasma')
```

and investigate the following functions,

- sort()
- order()
- table()


There are also plenty of functions for *creating* data. For example, **rnorm()** will give you random numbers from the normal distribution.

#### Exercise

1. Investigate rnorm from the help files and create 10 data-points with a mean of 10 and standard deviation of 2
2. Run your code several times. What do you notice?
3. Type 'set.seed(123)' above your code. Now, run both lines together several times. What do you notice?

Other useful functions are **seq()** and **rep()**, which general a sequence of numbers and repeated numbers (or words), respectively.

#### Exercise

1. Create a vector of numbers from 1 to 10
2. Create a second vector of the square root of the first vector
3. Pick 4 random values from the second vector (hint: use the 'sample()' function)


### Data types and structures

Data types in R mainly consist of the following,

- numeric
- integer
- character
- logical

The type of data you're dealing with will limit the sort of things that you can do with them, so it's important to know what you have. There are several ways to check, one of which is 'typeof()'.
We've already encountered each of these data types to some extent. The 'logical' data type is extremely important when dealing with missing values, and a very useful function when checking for
missing values is **is.na()**. For example,

```{r}
na_example = c(2, 5, 7, 7, NA, 3, 10, NA, 9, 2)
is.na(na_example)
```

This tells you where the missing values are, which you can then use with your code (which we'll see later)


#### Exercise

Using 2 of the above functions, how could you check *how many* missing values you have?


Data structures are collections of the above, and we've already seen one of these, too (vectors). The other main ones that you're likely to encounter are,

- list
- matrix
- data-frame

Lists are like vectors, but they can contain a mixture of data types. Matrices are 2-dimensional structures used in various areas of science and statistics. The data-type that most 
resembles a spreadsheet is the data-frame, which we'll see in the next section on loading data.

### Sub-setting

Whatever your data type, often you'll want to subset the data in order to use certain parts. For example, you may want the first element of a vector, or the last 3 elements, or the first column from a 
matrix. You may even wish to subset a character vector and pick out, say, the first letter.

#### Exercise

Subsetting typically works by indicting the position or index of the data that you want.

1. Preceded by set.seed(123), create the following vector,

```{r}
numbers = rnorm(n=5)
```


2. Type 'numbers' to see your generated vector
3. Type 'numbers[3]'. What do you get?
4. Now try 'numbers[3:5]

#### Exercise

1. Create the following matrix,

```{r}
matrix_example = matrix(rnorm(n=50), nrow = 10, ncol = 5)
```

2. Type 'matrix_example[,c(1:2)]'. What happens?
3. Type 'matrix_example[c(1:2),]'. What happens?


#### Exercise

1. Create the following character object,

```{r}
patient = 'Mrs. C. Ode
```

2. Type 'substr(patient, 1,3)'. What happens?

Subsetting not only works by specifying index values, but it can also be done based upon logical values. Effectively, picking out rows, columns or cells that are TRUE or FALSE. Earlier we created a vector called na_example. Let's see how you could *impute* the missing values using this subsetting idea,

```{r}
imputed = mean(example, na.rm = TRUE) #Determine the imputed value based upon the mean values
example[is.na(example)] = imputed #Set the missing values to this mean value
```

Finally, note that there are comments in the code, ignored by R thanks to the # character. Always consider using comments in your code to help interpret your code at a later date, or to help others to understand your code.