---
title: "Stats R Notebook Exercise"
output:
  html_notebook
---

Load the data,

```{r}
setwd('C:/Users/rob.harrand/OneDrive - Avacta Group plc/ANIMAL HEALTH/R Projects/life-sciences-workshop-dev-zone')
batch1 = read.csv('stats_example_batch1.csv')
batch2 = read.csv('stats_example_batch2.csv')
```

Check the mean difference,

```{r}
(mean(batch1$x) - mean(batch2$x)) / mean(batch2$x)
```

This is within limits (<10%). The new reagent is OK.

Visualise with a histogram,

```{r}
hist(batch1$x)
hist(batch2$x, col = rgb(1,0,0,alpha=0.25), add = T)
```

Perform a t-test,

```{r}
t.test(batch1$x, batch2$x, paired = TRUE)
```

Statistically, there is no difference (albeit just). Again, this is a pass for the new reagent.

Try batches 3 and 4,

```{r}
setwd('C:/Users/rob.harrand/OneDrive - Avacta Group plc/ANIMAL HEALTH/R Projects/life-sciences-workshop-dev-zone')
batch3 = read.csv('stats_example_batch3.csv')
batch4 = read.csv('stats_example_batch4.csv')
```

```{r}
(mean(batch3$x) - mean(batch4$x)) / mean(batch3$x)
```

This is again within limits. Another pass.

Visualise,

```{r}
hist(batch3$x, xlim = c(50,60))
hist(batch4$x, col = 'red', add = T)
```

This looks very different. t-test,

```{r}
t.test(batch3$x, batch4$x, paired = TRUE)
```

Statistically, the two batches are different.

What's going on here?

```{r}
mean(batch1$x)
sd(batch1$x)

mean(batch2$x)
sd(batch2$x)
```
```{r}
mean(batch3$x)
sd(batch3$x)

mean(batch4$x)
sd(batch4$x)
```






